Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data

### 2013-09-30 text/final_paper/ms.tex
final paper

\title{Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: A Statistical Search Through Palomar Transient Factory Data}
\author{Adrian~M.~Price-Whelan\altaffilmark{1,8} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda P. Fournier\altaffilmark{2,3}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{6}, Jason Surace\altaffilmark{7}}
%, Joshua S.\ Bloom\altaffilmark{\ref{cal}}, S.\ Bradley Cenko\altaffilmark{\ref{cal}}, Mansi M.\ Kasliwal\altaffilmark{\ref{calt}}, Shrinivas R.\ Kulkarni\altaffilmark{\ref{calt}}, Nicholas~M.~Law\altaffilmark{\ref{to},\ref{dun}}, Peter Nugent\altaffilmark{\ref{lbnl}}, Dovi Poznanski\altaffilmark{\ref{cal},\ref{calt},\ref{eins}}, Robert M.\ Quimby\altaffilmark{\ref{calt}}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{6}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}
%\altaffiltext{9}{Infrared Processing and Analysis Center, California Institute of Technology, Mail Stop 100-22, Pasadena, CA 91125, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{cal} Department of Astronomy, University of California, Berkeley, CA  94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{to} Dunlap Institute for Astronomy and Astrophysics, University of Toronto, 50 St.\ George Street, Toronto M5S 3H4, Ontario, Canada}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{lbnl} Computational Cosmology Center, Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
%\altaffiltext{\theaddress}{\stepcounter{address}\label{dun} Dunlap Fellow}
\altaffiltext{8}{NSF Graduate Fellow}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations, a total of $1.1\times10^9$ light curves, to uncover three candidate sparse-field microlensing events. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope.

% Second to last sentence will change when I run on all fields with >25 observations
\end{abstract}

\keywords{
  survey science
  ---
  gravitational microlensing
  ---
  time-domain astrophysics
}

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and, most recently, to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

%The microlensing event rate for a given region of the sky depends on the space and velocity distributions of the sources and on the space, velocity, and mass distribution of the lenses.
To maximize the event rate, microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.][]{original_ogle, original_macho, eros_original, crotts1996}. Of course, microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the event rate in low-density stellar regions will certainly be lower than that near, e.g., the Bulge \citep[][]{ogle_optical_depth, macho_optical_depth, wood_optical_depth, eros_optical_depth}, these events are expected to involve closer sources and lenses, and less crowded backgrounds, compared to those found by modern microlensing surveys \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

\begin{figure*}[!t]
\centering\includegraphics[scale=0.4,trim=0 0 0 27, clip]{figures/R_coverage_crop.pdf}
\caption{PTF $R$-band survey footprint, in equatorial coordinates. The fields are color-coded by the total number of observations. The field size corresponds to the actual area covered by a single PTF exposure.}\label{fig:survey_footprint}
\end{figure*}

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and, therefore, mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii, therefore, tend to be larger (a few milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would, therefore, be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, which is described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. A simple extrapolation from previous all-sky rate estimates suggests $\sim$1--10 detectable microlensing events within the PTF footprint over the three year survey \citep{han2008, gaudi2008}. We examine the selection methods used by microlensing surveys to identify event signatures in light curves and an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search}, we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}. A detailed investigation of the all-sky microlensing event rate,  generated from a model that includes realistic mass and velocity distributions, will be presented in Fournier et al. (in prep).

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48-nch Oschin Schmidt Telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (2.2$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2009,rau2009,nick2010}. %The data are processed in real time by a pipeline that i

While a real-time image-differencing pipeline identifies transients of interest and passes these to a dedicated photometric follow-up telescope, a separate pipeline generates PTF light curves. Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al., in prep). Catalogs are produced using SExtractor for source identification \citep{bertin96}. The absolute photometric calibration of the images is good to a systematic limit of $\sim$2\% for photometric nights, as described in \cite{ofek2012}.  The final light curves are produced using relative photometric calibration, which refines the calibration to $<$1\% for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{levitan2011, ofek2012}. The resulting archive is an excellent resource for the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}. %A separate pipeline identifies transients of interest and passes these to a dedicated photometric follow-up telescope; this has proved an excellent tool for e.g., early detection of supernovae \citep{nugent2011}.

%In addition, all detected sources are archived\footnote{The first photometric catalog of non-variable sources drawn from this archive is described in \citet[][]{ofek2012}.};

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.47\textwidth, trim=2 2 5 5, clip]{figures/sampling_figure.pdf}
\caption{Sampling of six different PTF fields over the course of a year. Darker points indicate that more exposures were taken that night; the total number of observations for each field is indicated.}\label{fig:sampling}
\end{center}
\end{figure}

\begin{figure}[t]
\centering\includegraphics[width=0.5\textwidth, trim=21 20 25 70, clip]{figures/num_observations_baseline_90deg.pdf}
\caption{Total number of $R$-band exposures and baselines (number of days between first and last exposure) for all PTF fields. The histogram to the right shows that the exposure distribution is peaked at 30-40; the histogram at the top shows that the baseline distribution has three peaks, corresponding to fields observed over $\lapprox$weeks, $\sim$1 year, and $\sim$3 years (the full length of the survey).}
\label{fig:num_obs}
\end{figure}

\begin{figure}[!h]
\centering\includegraphics[width=0.48\textwidth, trim=5 5 5 7, clip]{figures/simulated_events.pdf}
\caption{Simulating the effect of $20$-day microlensing events with impact parameters $u_0 = 1.0$, 0.5, and 0.01, on a random PTF $R$-band light curve. The original light curve is shown at the top. The dashed line in the bottom panel shows the approximate saturation limit of the PTF camera in the $R$-band; such an event with a brighter source might therefore be missed by the survey.}\label{fig:microlensing_sim}
\end{figure}

@@

In Figure~\ref{fig:num_obs}, we plot the number of $R$-band exposures for each field against the observational baseline (the number of days between the first and last exposure). Given the exposure distribution, which peaks at 30-40 but includes a long tail to larger numbers, we restrict our sample to the light curves with $>$10 \textit{R}-band observations. % because there is a drop off in the number of fields with fewer than 10 observations and we don't expect such light curves to be useful for our search. %APW!

@@

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$, but different angular impact parameter, $u_0$, on a random well-sampled PTF light curve.

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. The light curves of transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from, e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive data points brighter than some threshold,
	\item compared to flat or linear light-curve models, a microlensing model best describe the data, as determined by using a $\Delta\chi^2$ test, and
	\item the microlensing model event parameters have physically reasonable values.
\end{enumerate}
%Difference imaging is ideal for studying crowded fields where photometric measurements are difficult. It also limits the number of light curves that must be fit with a model, thus saving on computation time and allowing for real-time detection of events, so that high-cadence follow-up observations of interesting sources can be triggered.

@@

\subsection{The Variability Indices} %\label{sec:event_selection}
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$\footnote{We do not implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.}. $\sigma/\mu$ is the ratio of the sample standard deviation to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

\begin{figure}
\centering\includegraphics[width=.48\textwidth, trim=130 85 90 65, clip]{figures/max_outlier_light_curves.pdf}
\caption{The maximally outlying light curves for each variability index applied to data from PTF field 100101. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices, these are the light curves with the \emph{maximum} value of the relevant index over the field. } \label{fig:indices_examples}
\end{figure}

@@

\begin{figure}[t]
\centering
        \subfigure{
	   \includegraphics[width=.467\textwidth, trim=5 9 5 3, clip] {figures/eta_vs_delta_chi_squared.pdf}
	 }

	 \subfigure{
	   \includegraphics[width=.47\textwidth, trim=10 9 0 7, clip] {figures/eta_vs_j.pdf}
	 }

	 \subfigure{
	   \includegraphics[width=.47\textwidth, trim=10 6 0 7, clip] {figures/delta_chi_squared_vs_j.pdf}
	 }
	\caption{Two-dimensional density histograms for projections of the six-dimensional variability statistic distribution for 10,000 light curves from PTF field 100018 (left) and with the addition of simulated microlensing events (right). Red (dashed) lines are the 1\% false positive recovery selection boundaries for each index.} 
	\label{fig:var_indices}
\end{figure}

@@

@@

\begin{figure}
\centering
	\subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 3, clip] {figures/example_light_curve_f4327_ccd2.pdf}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {figures/field004327_Nperccd1000_Nevents100.pdf}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$, as a function of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$ (the event timescale, impact parameter, and `quiescent' source magnitude, respectively). Black (dashed) lines show the normalized input distributions, red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

\begin{figure}
\centering
        \subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 3, clip] {figures/example_light_curve_f4588_ccd2.pdf}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {figures/field004588_Nperccd1000_Nevents100.pdf}
	   \label{fig:subfig2}
	 }
\caption{Same as \figurename~\ref{fig:detection_efficiency_4327}, for PTF field 4588.}\label{fig:detection_efficiency_4588}
\end{figure}

\begin{figure}
\centering
        \subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 5, clip] {figures/example_light_curve_f100152_ccd2.pdf}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {figures/field100152_Nperccd1000_Nevents100.pdf}
	   \label{fig:subfig2}
	 }
\caption{Same as \figurename~\ref{fig:detection_efficiency_4327}, for PTF field 100152.}\label{fig:detection_efficiency_100152}
\end{figure}

@@

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 7 photometry \citep{DR7paper} to identify candidate quasars \citep[using the cuts described in][]{richards02} and examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by e.g., supernovae.

We visually inspect each of the remaining $\sim$2000 light curves. We identify an additional $\sim$1100 objects with unknown long-term variability (e.g, Mira-type variables, quasars not in the SDSS footprint). An additional $\sim$600 light curves have bad or poorly calibrated data that mimics a transient increase in flux: diffraction spikes, ghosts, and scattered light can cause non-Gaussian, seeing-dependent variability. Where appropriate, as an additional test, we examine the PTF images to verify the data quality. A small fraction of the selected light curves are variable stars that made it past our periodic variability cut.

>>>>

\title{Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data}
\author{Adrian~M.~Price-Whelan\altaffilmark{1,8} Marcel~A.~Ag\"ueros\altaffilmark{1}, Amanda P. Fournier\altaffilmark{2,3}, Rachel Street\altaffilmark{3}, Eran O. Ofek\altaffilmark{4}, Kevin~R.~Covey\altaffilmark{5}, David Levitan\altaffilmark{6}, Russ R.~Laher\altaffilmark{7}, Branimir Sesar\altaffilmark{6}, Jason Surace\altaffilmark{7}}

\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{Department of Physics, Broida Hall, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{3}{Las Cumbres Observatory Global Telescope Network, Inc., 6740 Cortona Dr.\ Suite 102, Santa Barbara, CA 93117, USA}
\altaffiltext{4}{Benoziyo Center for Astrophysics, Weizmann Institute of Science, 76100 Rehovot, Israel}
\altaffiltext{5}{Lowell Observatory, 1400 West Mars Hill Rd, Flagstaff, AZ 86001, USA}
\altaffiltext{6}{Division of Physics, Mathematics, and Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
\altaffiltext{7}{Spitzer Science Center, California Institute of Technology, Mail Stop 314-6, Pasadena, CA 91125, USA}

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal time sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events in our data. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations --- a total of $1.1\times10^9$ light curves --- to uncover three candidate sparse-field microlensing events. While these candidates cannot be confirmed, as we lack simultaneous, multi-color photometry of the events, this number is consistent with simple predictions for the microlensing event rate in the PTF footprint over the survey's three years of operations. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope. 
\end{abstract}

\keywords{
  gravitational lensing: micro
  ---
  surveys
  ---
  methods: statistical
}

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

Microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.,][]{original_ogle, original_macho, eros_original, crotts1996}, but microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the microlensing event rate in low-density stellar regions will be smaller, compared to events found by modern microlensing surveys, these events are expected to involve closer sources and lenses, and less crowded backgrounds \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

\begin{figure*}[!t]
\centering\includegraphics[scale=0.4,trim=0 0 0 27, clip]{fig1.eps}
\caption{PTF $R$-band survey footprint, in equatorial coordinates. The fields are color-coded by the total number of observations. The field size corresponds to the actual area covered by a single PTF exposure.}\label{fig:survey_footprint}
\end{figure*}

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and, therefore, mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii, therefore, tend to be larger ($\sim$milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would, therefore, be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. Although to date no microlensing event has been reported using PTF data, a simple extrapolation from previous all-sky rate estimates suggests $\sim$1--10 detectable microlensing events within the PTF footprint over the survey's three years of operation \citep{han2008, gaudi2008}. We examine the selection methods used by microlensing surveys to identify event signatures in light curves as well as an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search}, we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}. A detailed investigation of the all-sky microlensing event rate,  generated from a model that includes realistic mass and velocity distributions, will be presented in Fournier et al. (in prep).

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48-nch Oschin Schmidt Telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (2.2$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2009,rau2009,nick2010}.

While a real-time image-differencing pipeline identifies transients of interest and passes these to a dedicated photometric follow-up telescope, a separate pipeline generates PTF light curves. Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al., in prep). SExtractor is used for source identification \citep{bertin96}. The absolute photometric calibration of the images is good to a systematic limit of $\sim$2\% for photometric nights, as described in \cite{ofek2012}.  The final light curves are produced using relative photometric calibration, which refines the calibration to $<$1\% for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{levitan2011, ofek2012}. The resulting archive is an excellent resource for the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.47\textwidth, trim=2 2 5 5, clip]{fig2.eps}
\caption{Sampling of six different PTF fields over the course of a year. Darker points indicate that more exposures were taken that night; the total number of observations for each field is indicated.}\label{fig:sampling}
\end{center}
\end{figure}

\begin{figure}[t]
\centering\includegraphics[width=0.5\textwidth, trim=21 20 25 70, clip]{fig3.eps}
\caption{Total number of $R$-band exposures and baselines (number of days between first and last exposure) for all PTF fields. The histogram to the right shows that the exposure distribution is peaked at 30-40; the histogram at the top shows that the baseline distribution has three peaks, corresponding to fields observed over $\lapprox$weeks, $\sim$1 year, and $\sim$3 years (the full length of the survey).}
\label{fig:num_obs}
\end{figure}

\begin{figure}[!h]
\centering\includegraphics[width=0.48\textwidth, trim=5 0 0 7, clip]{fig4.eps}
\caption{Simulating the effect of $20$-day microlensing events with impact parameters $u_0 = 1.0$, 0.5, and 0.01, on a random PTF $R$-band light curve. The original light curve is shown at the top. The dashed line in the bottom panel shows the approximate saturation limit of the PTF camera in the $R$-band; such an event with a brighter source might therefore be missed by the survey.}\label{fig:microlensing_sim}
\end{figure}

In Figure~\ref{fig:num_obs}, we plot the number of $R$-band exposures for each field against the observational baseline (the number of days between the first and last exposure). Given the exposure distribution, which peaks at 30-40 but includes a long tail to larger numbers, we restrict our sample to the light curves with $>$10 \textit{R}-band observations.

@@

Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$, but different angular impact parameter, $u_0$, on a random well-sampled PTF light curve.

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. The light curves of transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from, e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive data points brighter than some threshold,
	\item compared to flat or linear light-curve models, a microlensing model best describe the data, as determined by using a $\Delta\chi^2$ test, and
	\item the microlensing model event parameters have physically reasonable values.
\end{enumerate}

@@

\begin{figure}[h]
\centering\includegraphics[width=.48\textwidth, trim=130 85 90 65, clip]{fig5.eps}
\caption{The maximally outlying light curves for each variability index applied to data from PTF field 100101. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices, these are the light curves with the \emph{maximum} value of the relevant index over the field. } \label{fig:indices_examples}
\end{figure}

\subsection{The Variability Indices} 
We choose five statistical measures of variability compiled by \cite{shin2009} that have been previously applied to the classification and discovery of periodic variables. These variability indices are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$\footnote{We do not implement the sixth index described by \cite{shin2009}, $AoVM$, because it mainly helps identify periodic sources.}. $\sigma/\mu$ is the ratio of the sample standard deviation to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations.

@@

\begin{figure}[h]
\centering
        \subfigure{
	   \includegraphics[width=.467\textwidth, trim=5 9 5 3, clip] {fig6_a.eps}
	 }

	 \subfigure{
	   \includegraphics[width=.47\textwidth, trim=10 9 0 7, clip] {fig6_b.eps}
	 }

	 \subfigure{
	   \includegraphics[width=.47\textwidth, trim=10 6 0 7, clip] {fig6_c.eps}
	 }
	\caption{Two-dimensional density histograms for projections of the six-dimensional variability statistic distribution for 10,000 light curves from PTF field 100018 (left) and with the addition of simulated microlensing events (right). Red (dashed) lines are the 1\% false positive recovery selection boundaries for each index. Shaded (red) regions show the subspaces where microlensing events are expected to fall, as defined by the two statistics in question. } 
	\label{fig:var_indices}
\end{figure}

@@

@@

\begin{figure}
\centering
	\subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 3, clip] {fig7_a.eps}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {fig7_b.eps}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$, as a function of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$ (the event timescale, impact parameter, and `quiescent' source magnitude, respectively). Black (dashed) lines show the normalized input distributions, red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

@@

\begin{figure}
\centering
        \subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 3, clip] {fig8_a.eps}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {fig8_b.eps}
	   \label{fig:subfig2}
	 }
\caption{Same as \figurename~\ref{fig:detection_efficiency_4327}, for PTF field 4588.}\label{fig:detection_efficiency_4588}
\end{figure}

\begin{figure}
\centering
        \subfigure{
	   \includegraphics[width=.45\textwidth, trim=0 8 6 5, clip] {fig9_a.eps}
	   \label{fig:subfig1}
	 }

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {fig9_b.eps}
	   \label{fig:subfig2}
	 }
\caption{Same as \figurename~\ref{fig:detection_efficiency_4327}, for PTF field 100152.}\label{fig:detection_efficiency_100152}
\end{figure}

@@

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 7 photometry \citep{DR7paper} to identify candidate quasars \citep[using the cuts described in][]{richards02} and examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by, e.g., supernovae.

\begin{figure*}[htb]
\centering
%\includegraphics[width=2in] {PTFS1206i_lc.png}%}
%\hspace{.2cm}
%\includegraphics[width=2in, trim = 0 0 -30 0, clip] {PTFS1216bt_lc.png}%}
%\hspace{.2cm}
%\includegraphics[width=2in] {PTFS1217cv_lc.png}%}
\includegraphics[width=2.1\columnwidth]{fig10.eps}
\caption{Full light curves (top) and zooms around the transient maximum (bottom) for the three plausible microlensing event candidates. Bad (flagged) data points have been removed. Lines (gray, thin) show models with parameters sampled from the posterior probability distribution over the four parameters in the point-lens, point-source microlensing event model. Black (thick) line shows the maximum {\it a posteriori} model. }\label{fig:candidates}
\end{figure*}
